---
title: "Assignment 6: GLMs (Linear Regressios, ANOVA, & t-tests)"
author: "Maeve Arthur"
date: "Spring 2023"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on generalized linear models. 

## Directions
1. Rename this file `<FirstLast>_A06_GLMs.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up your session 
1. Set up your session. Check your working directory. Load the tidyverse, agricolae and other needed packages. Import the *raw* NTL-LTER raw data file for chemistry/physics (`NTL-LTER_Lake_ChemistryPhysics_Raw.csv`). Set date columns to date objects.

2. Build a ggplot theme and set it as your default theme.

```{r setup2}
#1
library(tidyverse)
library(agricolae)   
library(here)
library(lubridate)
here()

Lakes <- read.csv(here("Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv"), stringsAsFactors = TRUE)

# Set date to date format
Lakes$sampledate <- mdy(Lakes$sampledate); class(Lakes$sampledate)
#should we change year column too?

#2 making theme
mytheme <- theme_classic(base_size = 14) +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "top")
theme_set(mytheme)

```

## Simple regression
Our first research question is: Does mean lake temperature recorded during July change with depth across all lakes?

3. State the null and alternative hypotheses for this question:
> Answer:
H0: Mean lake temperature in July does not change with depth across across all lakes.
Ha: Mean lake temperature in July does change with depth across all lakes.

4.  Wrangle your NTL-LTER dataset with a pipe function so that the records meet the following criteria: 
 * Only dates in July. 
 * Only the columns: `lakename`, `year4`, `daynum`, `depth`, `temperature_C`
 * Only complete cases (i.e., remove NAs)

5. Visualize the relationship among the two continuous variables with a scatter plot of temperature by depth. Add a smoothed line showing the linear model, and limit temperature values from 0 to 35 °C. Make this plot look pretty and easy to read.

```{r scatterplot}
#4 wrangling
Lakes.wrangled <- 
  Lakes %>%
  filter(month(sampledate) == 7) %>%
  select(c('lakename', 'year4', 'daynum', 'depth', 'temperature_C')) %>%
  na.omit()

#5 scatterplot
temp.by.depth <- 
  ggplot(Lakes.wrangled, aes(x = depth, y = temperature_C)) +
  geom_point() +
  geom_smooth(method="lm", col="blue", se=FALSE) +
  ylim(0,35) +
  labs(x="Depth (m)", y = "Temperature (C)", title = "Lake Temperature by Depth in July")
  
print(temp.by.depth) 

```


6. Interpret the figure. What does it suggest with regards to the response of temperature to depth? Do the distribution of points suggest about anything about the linearity of this trend?

> Answer: The plot suggests that temperature increases as depth decreases. The distribution of points suggest that this relatioinship is not fully linear. There are points both above and below the line.


7. Perform a linear regression to test the relationship and display the results

```{r linear.regression}
#7 linear regression through lm function
temp.depth.regression <- lm(Lakes.wrangled$temperature_C ~ Lakes.wrangled$depth)
summary(temp.depth.regression)

```


8. Interpret your model results in words. Include how much of the variability in temperature is explained by changes in depth, the degrees of freedom on which this finding is based, and the statistical significance of the result. Also mention how much temperature is predicted to change for every 1m change in depth. 

> Answer: The model shows a p-value of < 2.2e-16, which less than .05, meaning that changes in temperature can be explained by changes in depth. The model also shows an r-sqaured value of .7387, meaning that 73.87% of the variability in temperature is explained by changes in depth. These findings are based on 9,726 degrees of freedom.  For every 1m change in depth, temperature is predicted to change by -1.94621 degrees celcius.
---

## Multiple regression
Let's tackle a similar question from a different approach. Here, we want to explore what might the best set of predictors for lake temperature in July across the monitoring period at the North Temperate Lakes LTER. 


9. Run an AIC to determine what set of explanatory variables (year4, daynum, depth) is best suited to predict temperature.

10. Run a multiple regression on the recommended set of variables. 

```{r temperature.model}
#9 lake AIC
LakeAIC <- lm(data=Lakes.wrangled, temperature_C~depth + year4 + daynum)
summary(LakeAIC)
step(LakeAIC)

#10 multiple regresion
Lake.best <- lm(data=Lakes.wrangled, temperature_C~depth + year4 + daynum)
summary(Lake.best)

```

11. What is the final set of explanatory variables that the AIC method suggests we use to predict temperature in our multiple regression? How much of the observed variance does this model explain? Is this an improvement over the model using only depth as the explanatory variable?

> Answer: The AIC suggested to use depth, year4, and daynum because all of them are significant explanatory variables. 74.11% of the variance is explained by this model. This is a slight improvement over only using depth, which eplained 73.87% of the variability. They are very similar.

---
## Analysis of Variance

12. Now we want to see whether the different lakes have, on average, different temperatures in the month of July. Run an ANOVA test to complete this analysis. (No need to test assumptions of normality or similar variances.) Create two sets of models: one expressed as an ANOVA models and another expressed as a linear model (as done in our lessons).

```{r anova.model}
#12
# Format ANOVA as aov
Lake.anova <- aov(data = Lakes.wrangled, temperature_C~lakename)
summary(Lake.anova)

# Format ANOVA as lm 
Lake.anova2 <- lm(data = Lakes.wrangled, temperature_C~lakename)
summary(Lake.anova2)

```

13. Is there a significant difference in mean temperature among the lakes? Report your findings. 

> Answer: The models showed a p-value of p-value: < 2.2e-16. Since this is less than 0.05, there is a significant differrence in mean temperature among the lakes.


14. Create a graph that depicts temperature by depth, with a separate color for each lake. Add a geom_smooth (method = "lm", se = FALSE) for each lake. Make your points 50 % transparent. Adjust your y axis limits to go from 0 to 35 degrees. Clean up your graph to make it pretty. 

```{r scatterplot.2}
#14. temp by depth scatterplot with color by lakename
temp_by_depth_2 <- 
  ggplot(Lakes.wrangled, aes(x = depth, y = temperature_C, col=lakename)) +
  geom_point(alpha=0.5) +
  geom_smooth(method="lm", se=FALSE) +
  ylim(0,35) +
  labs(
    x="Depth (m)", 
    y = "Temperature (C)",
    title = "Lake Temperature by Depth and Lakename in July",
    color = "Lake Name") +
  theme(
    legend.text = element_text(size = 8),
    legend.position = "right")
  
print(temp_by_depth_2) 


```

15. Use the Tukey's HSD test to determine which lakes have different means.

```{r tukey.test}
#15 tukey test
TukeyHSD(Lake.anova)

Lake.groups <- HSD.test(Lake.anova, "lakename", group = TRUE)
Lake.groups

```

16.From the findings above, which lakes have the same mean temperature, statistically speaking, as Peter Lake? Does any lake have a mean temperature that is statistically distinct from all the other lakes?

>Answer: The lakes that have the same mean temperature as Peter Lake are Peter Paul and Ward Lake. There is not one lake that has a mean temperature that is statistically distinct.


17. If we were just looking at Peter Lake and Paul Lake. What's another test we might explore to see whether they have distinct mean temperatures? 

>Answer: We would use a two-sample t-test to see if Peter Lake and Paul Lake have distinct mean temperatures.

18. Wrangle the July data to include only records for Crampton Lake and Ward Lake. Run the two-sample T-test on these data to determine whether their July temperature are same or different. What does the test say? Are the mean temperatures for the lakes equal? Does that match you answer for part 16?

```{r t.test}
#wrangling data
Lakes.wrangled.2 <- 
Lakes.wrangled %>%
  filter(lakename == "Crampton Lake" | lakename == "Ward Lake")

#two sample t-test
Lake.twosample <- t.test(Lakes.wrangled.2$temperature_C ~ Lakes.wrangled.2$lakename)
Lake.twosample

```

>Answer: The t-test reveals a p-value of 0.2649, meaning that the means in Crampton lake and Ward Lake are not significantly different. This confirms my answer from part 16.
